# Preparing Data for Analysis {#sec-data_munging}

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

In [Chapter @sec-intro_tidyverse], we saw the main functions of the `tidyverse`. In this chapter, we will explore advanced uses of these, as well as introduce functions for merging and reshaping `tibbles`, key steps in LDA. The aim is to make analysis-ready datasets in as efficient a manner as possible.

The packages we will use are:

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(haven)
library(glue)
library(labelled)
```

We'll begin using data from the MCS Sweep 7 (17y) cohort member interview dataset.

```{r}
mcs_fld <- Sys.getenv("mcs_fld")
ncds_fld <- Sys.getenv("ncds_fld")

mcs_17y <- glue("{mcs_fld}/17y/mcs7_cm_interview.dta") %>%
  read_dta()
```

## Advanced Functions for Data Cleaning

### `tidyselect` Helpers

In [Chapter @sec-intro_tidyverse], we used `select()` to retain or drop variables from a `tibble`. We saw the manual way of doing this (specifying each variable separately), as well as a more programmatic way: using a `tidyselect` helper, `matches()`, that selected every variable whose name matched a regular expression. `matches()` is one of many [`tidyselect` helpers](https://tidyselect.r-lib.org/reference/language.html):

-   `matches()`: selects variables based on regular expressions (e.g., `matches("^G")` selects any variable starting `"G"`).
-   `x:y`: selects variables location between `x` and `y` (inclusive). `x` and `y` can be variable names and/or positions. `last_col()` is another useful helper that can be used in this context (e.g., `age:last_col()` selects every column from `age` to the last column).
-   `where()`: selects variables based on a logical condition - this condition is applied to the values not the variable names. For example, `where(is.factor)` selects all factor variables.
-   `all_of()`: selects all variables specified in a character vector (e.g., `all_of(c("sex", "age")` selects columns `sex` and `age`). Note, `all_of()` throws an error if not all of the specified variables are present in the `tibble`.
-   `any_of()`: returns any variables specified in a character vector that are available in a `tibble`. Unlike `all_of()`, `any_of()` does not throw an error if a specified variable is not present.
-   `everything()`: selects every column in a `tibble`. This may seem redundant, but we will see contexts in which it is useful.

`tidyselect` helpers can be combined with the `&` (`AND`) and `|` (`OR`) operators, while the `!` (`NOT`) operator can be used to identify variables that do not meet a condition. For example, `where(is.numeric) & !matches("^G")` selects all numeric variables that do not start `"G"`. Below is some example code.

> **Question**: Which variables will remain in the outputted `tibble`?

```{r}
mcs_17y %>%
  select(matches("^GCW")) %>%
  select(GCWESM00:last_col()) %>%
  select(where(is.labelled)) %>%
  select(all_of(c("GCWWOP00", "GCWWUS00", "GCWWRE00"))) %>%
  select(any_of(c("GCWWOP00", "GCWWRE00", "not_a_variable"))) %>%
  select(!matches("^GCWWO"))
```

`tidyselect` helpers can be used in many functions in the `tidyverse` beyond `select()`. For instance, they can be used to specify columns in `drop_na()`, in the `col_select = ...` argument of `read_dta()`, the `.cols = ...` argument of `rename_with()`, and in `across()` (more on which below).

`tidyselect` helpers are not just useful for selecting raw variables en masse, but can also be used with cleaned variables too. This is made easier if care is taken to name cleaned variables consistently. For instance, naming BMI variables in a 'wide' dataset in a consistent way (e.g., `bmi_03`, `bmi_05`, ..., `bmi_17`) allows one to use `matches("^bmi_\\d\\d$")` to select them. Later, we will use vectors that specify control variables to be used in regression models and select them with `all_of()`.

### `dplyr::rename_with()`

Variables can be renamed en masse using `rename_with()`. `rename_with()` applies a function to each variable name in turn. Functions can be specified using the lambda notation we used in [Chapter @sec-intro_tidyverse] with `purrr:map()`. The [`stringr` package](https://stringr.tidyverse.org/) contains several functions that can be used inside `rename_with`, including `str_replace()` (to replace part of a string with another string), `str_remove()` (to remove part of a string), `str_to_lower()` (to put strings in lower case), and `str_sub()` (to extract characters between specified positions in a string; e.g., `str_sub(x, 2, -2)` extracts the second to penultimate letters of `x`). `glue::glue()` can also be handy. Here are some examples:

```{r}
#| output: false

mcs_17y %>%
  rename_with(str_to_lower) %>% # Equivalent to ~ str_to_lower(.x)
  rename_with(~ str_remove(.x, "^g")) %>%  # Removes sweep-specific prefix "G"
  rename_with(~ glue("{.x}_17"),
              .cols = -c(mcsid, cnum00))
```

Notice the line `rename_with(str_to_lower)` - when just a function name is specified, the column name will be placed in the first argument of that function. This syntax can also be used in `purrr::map()`.

`rename_with(~ str_remove(.x, "^G"))` is an essential preparatory step if one wishes to append data from different sweeps of the MCS together.

### Task I

1.  Open the age 50y and 62y main interview files (`ncds_2008_followup.dta` and `ncds10_age62_main_interview.dta`, respectively). Use `tidyselect` helpers to select every derived variable in these datasets.

```{r}
#| output: false
#| code-fold: true
#| code-summary: "Reveal the solution"

ncds_50y <- glue("{ncds_fld}/50y/ncds_2008_followup.dta") %>% 
  read_dta() 
ncds_62y <- glue("{ncds_fld}/62y/ncds10_age62_main_interview.dta") %>% 
  read_dta()

lookfor(ncds_50y, "derived")
lookfor(ncds_62y, "derived")

ncds_50y %>%
  select(NCDSID, matches("ND8"))

ncds_62y %>%
  select(ncdsid, matches("nd10"))
```

2.  Consider how to rename the variables in both datasets so that they could be appended together.

```{r}
#| output: false
#| code-fold: true
#| code-summary: "Reveal the solution"

names(ncds_50y)
names(ncds_62y)

harmonise_names <- function(df){
  df %>%
    rename_with(str_to_lower) %>%
    rename_with(~ str_remove(.x, "^nd(8|10)") %>%
                  paste0("_derived"),
                .cols = matches("^nd(8|10)")) %>%
    rename_with(~ str_remove(.x, "^n(8|10)"),
                .cols = -matches("_derived$"))
}

harmonise_names(ncds_50y)
harmonise_names(ncds_62y)
```

### `dplyr::across()`

`across()` allows one to apply a function to multiple variables in turn. It is most typically used in `mutate()` to transform multiple variables in the same way. A key use of `across()` in LDA is to replace negative values representing missing codes with `R`'s `NA` missing value. In [Chapter @sec-intro_tidyverse], we used `ifelse()` to recode negative values, but this drops variable metadata (e.g., value labels) that we may wish to keep. To recode missing values while retaining other metadata, we can use functionality from the `labelled` package as in the `negative_to_na()` function below, while we apply to all labelled variables (`where(is.labelled)`) using `across()`.

```{r}
negative_to_na <- function(x){
  na_range(x) <- c(-Inf, -1)
  user_na_to_na(x)
}

mcs_17y %>%
  select(MCSID, GCNUM00, matches("^GCHT(CM|FT|IN)00")) %>%
  mutate(across(where(is.labelled), ~ negative_to_na(.x)))
```

By default, `mutate(across())` overwrites the original variables. New columns can be created instead by specifying a named list of functions. The `.names` argument of `across()` allows one to also specify a rubric for the new columns names and uses `glue()`-like syntax.

```{r}
mcs_17y %>%
  select(matches("^GCHT(CM|FT|IN)00")) %>%
  mutate(across(where(is.labelled), ~ negative_to_na(.x))) %>%
  summarise(across(where(is.numeric),
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        sd = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.col}_{.fn}"))
```

`across()` appears in a few other places in the `tidyverse`, including `distinct()` and `count()`. If a `tidyselect` helper doesn't work on its own in a function, wrapping it in `across()` often does the trick.

```{r}
mcs_17y %>%
  mutate(across(where(is.labelled), ~ negative_to_na(.x))) %>%
  count(across(matches("^GCWW")))
```

### Some Helpful Functions for Labelled Data

Sometimes, in longitudinal datasets, some positive values can also indicate missingness, particularly in the context of a specific analysis. For instance, some self-complete items in the Sweep 7 of the MCS use positive values for responses that could be considered missing (e.g., "Don't know" or "Prefer not to say"). Specific values can be replaced with `NA` using the code below. This code also uses `haven::as_factor()`, which converts `labelled` vectors to factors using the value labels as levels.

```{r}
values_to_na <- function(x, values){
  na_values(x) <- values
  user_na_to_na(x)
}

mcs_17y %>%
  mutate(self_rated_health = negative_to_na(GCCGHE00) %>%
           values_to_na(6:8) %>%
           as_factor()) %>%
  count(GCCGHE00, self_rated_health)
```

After replacing missing value codes with `NA`, a `labelled` vector may not actually contain any value labels - e.g., where it is a continuous variable, like height. One operation you may wish to perform en masse then is to replace negative values with `NA` and to convert any variables with remaining value labels to factor variables or to remove the labelled class, otherwise, using `haven::zap_labels()`. This can be achieved as below:

```{r}
fix_labels <- function(x){
  if (is.null(val_labels(x))){
    clean <- zap_labels(x)
  } else{
    clean <- as_factor(x)
  }
  return(clean)
}

mcs_17y %>%
  select(GCSDQH00, GCHTCM00) %>%
  mutate(across(everything(), 
                ~ negative_to_na(.x) %>%
                  fix_labels()))
```

After cleaning data, it is good practice to add a descriptive label to each variable to make clear what each represents ([Vable et al., 2021](https://doi.org/10.1093/aje/kwab092)). The `labelled` package contains the `set_variable_labels()` function to do this. The typical syntax uses `column = "label"` syntax, but a named vector can instead be passed via the `.labels` argument. By default, setting the label for a variable that isn't present in the `tibble` will throw an error. This can be changed by setting the argument `.strict = FALSE` - this may be helpful when working with many datasets, each of which may not contain the exact same columns.

```{r}
mcs_17y %>%
  mutate(weight = negative_to_na(GCWTCM00),
         height = negative_to_na(GCHTCM00) / 100,
         bmi = weight / (height ^ 2)) %>%
  select(MCSID, cnum = GCNUM00, 
         height, weight, bmi) %>%
  set_variable_labels(
    MCSID = "Family ID",
    cnum = "Cohort Member Number",
    height = "Height (cm)",
    weight = "Weight (kg)",
    bmi = "Body Mass Index (kg/m^2)"
  )
```

### `dplyr::if_any()` / `dplyr::if_all()`

`if_any()` and `if_all()` are similar to `across()` in that they apply a function to multiple variables at once. They differ in that the function aggregates the results to return a logical vector. `if_any()` returns `TRUE` if the function returns `TRUE` for any of the variables, while `if_all()` returns `TRUE` if the function returns `TRUE` for all of the variables. `if_any()` and `if_all()` are most typically used in `filter()`, but can also be used in `mutate()`. Both `if_any()` and `if_all()` are vectorised.

```{r}
mcs_17y %>%
  mutate(across(where(is.labelled), ~ negative_to_na(.x))) %>%
  count(across(matches("^GCWW"))) %>%
  filter(if_any(matches("^GCWW"), ~ .x == 1))

mcs_17y %>%
  mutate(across(where(is.labelled), ~ negative_to_na(.x))) %>%
  count(across(matches("^GCWW"))) %>%
  filter(if_all(matches("^GCWW"), ~ .x == 1))
```

### `dplyr::pick()`

Sometimes we may want to perform a function using several columns simultaneously - for instance, for a mental health measure, to sum the values across several columns into a total Likert score. Base `R` contains several functions for carrying out row-wise operations, such as `pmin()`, `pmax()`, `rowSums()` and `rowMeans()`, but these functions are not by themselves pipe-friendly as they use `data.frames` or `tibbles` as inputs and operate over it all, which may be more columns than one wants, outputting a single vector. `pick()` allows one to specify the correct columns for these functions and place them within a pipe. In essence, `pick()` takes the current `tibble` and temporarily creates another which has just the columns specified. `tidyselect` helpers can be used to specify columns.[^05_data_preparation-1]

[^05_data_preparation-1]: Row-wise operations can also be carried out using `dplyr::rowwise()` and `dplyr::c_across()`. `rowwise()` is a special case of `group_by()`, and groups each row in a `tibble` separately. `c_across()` combine multiple columns together into a single object (e.g., a single vector), which may be necessary to use some functions. See the [`dplyr` article on row-wise operations](https://dplyr.tidyverse.org/articles/rowwise.html) for more detail.

```{r}
mcs_17y %>%
  select(matches("^GCWW")) %>%
  mutate(across(everything(), ~ negative_to_na(.x)),
         swemwbs_total = pick(matches("^GCWW")) %>% rowSums())
```

### Task II

1.  Open the Age 55y NCDS main interview file (`ncds_2013_flatfile.dta`). Create a CASP-6 Quality of Life score, where higher values indicate better quality of life. To construct this score take the mean item response and set to the score to missing if the cohort member has four or more items missing (hint: `rowSums(is.na(df)` calculate the number of missing values in a row). Be aware of how each item is coded when answering this question.

```{r}
#| output: false
#| code-fold: true
#| code-summary: "Reveal the solution"

ncds_55y <- glue("{ncds_fld}/55y/ncds_2013_flatfile.dta") %>% 
  read_dta()

lookfor(ncds_55y, "casp")

ncds_55y %>%
  mutate(across(matches("N9CASP"), negative_to_na)) %>%
  lookfor("casp")

ncds_55y %>%
  select(matches("N9CASP")) %>%
  mutate(across(matches("N9CASP"), negative_to_na),
         across(matches("N9CASP[4-6])"), ~ 5 - .x),
         casp6_mean = rowMeans(pick(matches("N9CASP[1-6]")), na.rm = TRUE), 
         casp6_missing = rowSums(is.na(pick(matches("N9CASP[1-6]")))),
         casp6_mean = ifelse(casp6_missing >= 4, NA, casp6_mean)) %>%
  group_by(casp6_missing) %>%
  slice(1) %>%
  ungroup()
```

XX-`naniar::n_miss_row()` - mention naniar is useful in general

2. Sweep 6 (14y) of the MCS included several cognitive assessments, including a 20-item vocabularly test. Answers to the individual items are contained in the variables `FCCMCOGA` to `FCCMCOGT` within the `mcs6_cm_cognitive_assessment.dta` file. The value label indicates which answer is correct. Load the data and create a variable for a cohort member's total score on the test as the number of correct items. Only give a score to an individual if they attempt one or more items.

```{r}
#| output: false
#| code-fold: true
#| code-summary: "Reveal the solution"

mcs_14y_cog <- glue("{mcs_fld}/14y/mcs6_cm_cognitive_assessment.dta") %>%
  read_dta()

lookfor(mcs_14y_cog)
lookfor(mcs_14y_cog, "FCCMCOG[A-T]")

val_labels(mcs_14y_cog$FCCMCOGA)
val_labels(mcs_14y_cog$FCCMCOGT)

clean_vocab_item <- function(item){
  value_labels <- val_labels(item)
  correct_answer <- value_labels[str_detect(names(value_labels), "\\(correct answer\\)")]
  
  case_when(item == correct_answer ~ 1,
            item %in% 1:5 ~ 0)
}

mcs_14y_cog %>%
  select(MCSID, FCNUM00, matches("FCCMCOG[A-T]")) %>%
  mutate(across(matches("FCCMCOG[A-T]"), clean_vocab_item),
         cog_score = pick(matches("FCCMCOG[A-T]")) %>% rowSums(na.rm = TRUE),
         cog_obs = pick(matches("FCCMCOG[A-T]")) %>% rowSums(!is.na(.)),
         cog_score = ifelse(cog_obs >= 1, cog_score, NA)) %>%
  select(MCSID, FCNUM00, cog_score)
```

## Writing Functions to Clean Data

As has been discussed, cleaning data is often a repetitive process in LDA: we may need to clean the same variables in the same way across different sweeps. In this case, to make our code efficient and less error-prone, we can write bespoke functions. Below is a function that cleans height and weight data and derives BMI in each of sweeps 3-7 in the MCS. As we saw in Chapter XX-FINDING, height and weight are contained in the variables `X-XX` and `X-XX` in each of these sweeps (where `X` is a letter denoting sweep). This makes the cleaning predictable.

```{r}
mcs_fups <- c(0, 3, 5, 7, 11, 14, 17)

clean_bmi <- function(sweep){
  sweep_letter <- letters[sweep]
  fup <- mcs_fups[sweep]
  glue("{mcs_fld}/{fup}y/mcs{sweep}_cm_interview.dta") %>%
    read_dta(col_select = c("MCSID",
                            matches("^.(CNUM|C(W|H)TCM)(A|0)0"))) %>%
    rename(cnum = matches(".CNUM00"),
           height_raw = matches(".CHTCM(A|0)0"),
           weight_raw = matches(".CWTCM(A|0)0")) %>%
    mutate(height = negative_to_na(height_raw) %>%
             as.numeric(),
           weight = negative_to_na(weight_raw) %>%
             as.numeric(),
           bmi = weight / ( (height / 100) ^ 2),
           fup = !!fup) %>%
    select(MCSID, cnum, fup, height, weight)
}

map(3:7, ~ clean_bmi(.x))
```

Notice the use of the `!!` (bang-bang) operator in `mutate(..., fup = !!fup)`. This is a form of ["quasiquotation"](https://adv-r.hadley.nz/quasiquotation.html) that allows one to use the value of an external object (e.g., `fup <- mcs_fups[sweep]`) in a `tibble` rather than the column (if they share the same name). We will return to this in \[Chapter @sec-regressions\].

## Combining `tibbles`

A key step in LDA is combining different datasets together - e.g., from different sweeps. This can be done by merging - placing the data in wide format - or by appending - placing the data in long format. We will start by discussing merging and then discuss appending.

To make the process easier to understand, we will use small datasets from sweeps 6 & 7, containing data on height (`(F|G)CHTCM00`) and weight (`(F|G)CWTCM00`) from three cohort members (`MCSIDs`: `M10020R`, `M10025W`, and `M10002P`) who appeared in sweeps 6 only, 7 only, and both sweeps, respectively. Later, to show how to combine more than two datasets at once, we will add in data from the same individuals from Sweep 5, which they each appeared in.

```{r}
load_mcs_mini <- function(sweep){
  sweep_letter <- letters[sweep]
  fup <- mcs_fups[sweep]
  
  glue("{mcs_fld}/{fup}y/mcs{sweep}_cm_interview.dta") %>%
    read_dta(col_select = c("MCSID", matches("^.(CNUM|C(W|H)TCM)(A|0)0"))) %>%
    filter(MCSID %in% c("M10020R", "M10025W", "M10002P")) %>%
    arrange(MCSID)
}

mcs_14y_mini <- load_mcs_mini(6)
mcs_14y_mini

mcs_17y_mini <- load_mcs_mini(7)
mcs_17y_mini
```

### Merging `tibbles` with `dplyr::*_join()`

To merge two `tibbles` together - i.e., to create a new `tibble` with columns from both `tibbles` - we can use the `dplyr` mutating `*_join()` functions. These are called mutating joins because they add new columns to the outputted `tibble`. The functions differ in the rows that each retain in the output `tibble`.

-   `full_join(x, y, by = ...)`: Returns all rows from `x` and `y` . The missing value `NA` is used to fill cells that are created during the merge. Rows in `x` with no match in `y` will have `NA` values in the new columns from `y`. Rows in `y` with no match in `x` will have `NA` values in the columns of `x`.
-   `inner_join(x, y, by = ...)`: Returns all rows from `x` and `y` where there are matching rows in both `tibbles`.
-   `left_join(x, y, by = ...)`: Returns all rows from `x` Rows in `x` with no match in `y` will have `NA` values in the new columns from `y`.
-   `right_join(x, y, by = ...)`: Returns all rows from `y`, and all columns from `x` and `y.` Rows in `y` with no match in `x` will have `NA` values in the columns of `x`.

To join the two datasets together, we need to specify the columns to join by (the "key" variables). By default, `*_join()` functions will use all variables with common names across the two datasets as the key variables. To specify particular columns, we can instead provide a character vector of the joining columns to the `by` argument. Joining columns do not need to have the same name in each `tibble`. If they have different names, we can specify a named vector. The names of this vector are names of the columns in the left (`x`) `tibble` and the values of the vector are the names of the columns in the right (`y`) `tibble`. In the merged `tibble`, key columns retain the names from the left `tibble`. Let's see the functions in action:

```{r}
full_join(mcs_14y_mini,
          mcs_17y_mini,
          by = c("MCSID", FCNUM00 = "GCNUM00"))

left_join(mcs_14y_mini,
          mcs_17y_mini,
          by = c("MCSID", FCNUM00 = "GCNUM00"))

right_join(mcs_14y_mini,
           mcs_17y_mini,
           by = c("MCSID", FCNUM00 = "GCNUM00"))

inner_join(mcs_14y_mini,
           mcs_17y_mini,
           by = c("MCSID", FCNUM00 = "GCNUM00"))
```

Given cohort members only appear in datasets if they have been surveyed in that sweep, rows retained after a specific `*_join()` imply a particular response pattern. For instance, `inner_join()` implies response in both sweeps.

> **Question**: what response patterns does appearing in the `tibble` outputted by `full_join()` imply? What about `left_join()` and `right_join()`?

More complicated rules for identifying rows to merge than above can be specified with the `join_by()` helper function. This is unlikely to be needed for most LDA applications, so we do not discuss this further here (an exception is joining overlapping activity history datasets) - see the [`dplyr` website](https://dplyr.tidyverse.org/reference/join_by.html) for more detail.

#### Merging 2+ `tibbles` with `purrr::reduce()`

What if we want to merge more than two datasets together - for instance, merge all seven sweeps of the MCS? Writing out each `*_join()` function separately would be cumbersome. Instead, we can use `purrr::reduce()` to chain together successive `*_join()` function calls, aggregating the result.

`reduce()` is similar to `map()` in that repeats a function for each object within an input. But unlike `map()`, `reduce()` uses two objects at a time, with the result fed back into the next function call. Specifically, `reduce()` takes an object, applies a function, `f`, to the first two element and then takes the result and inputs it into `f` along with the next element of the inputted object, repeating this until the object is object. For instance, specifying `inner_join()` in `reduce()` with a list of three `tibbles` will perform `inner_join()` on the first two, and then perform `inner_join()` on the result with the third `tibble`.

The syntax for `reduce()` is similar to `map2()` where `.x` and `.y` refer to the first and second elements of the list, respectively, or to the result and to the next element of the list). Below we perform an `inner_join()` on data from sweeps 5-7. Note, as the `XCNUM00` joining keys differ depending on which sweeps are being merged, we need to slightly alter the code used above to account for this.[^05_data_preparation-2]

[^05_data_preparation-2]: It is possible to seed `reduce()` with an initial `.x` value using the `.init` argument. This strategy can make sense for some transformations.

```{r}
mcs_11y_mini <- load_mcs_mini(5) 

list(mcs_11y_mini,
     mcs_14y_mini, 
     mcs_17y_mini) %>%
  reduce(
    ~ inner_join(
      .x %>% rename(CNUM = matches("CNUM")),
      .y %>% rename(CNUM = matches("CNUM")) ,
      by = c("MCSID", "CNUM")
    )
  )
```

> **Question**: what response pattern does appearing in the `tibble` outputted by `reduce(list(...), ~ inner_join(.x, .y))` imply? What would `reduce(list(...), ~ full_join(.x, .y))` and `reduce(list(...), ~ left_join(.x, .y)`) each imply?

#### Filtering Joins

`dplyr` also contains two filtering join functions, `semi_join(x, y)` and `anti_join(x, y)`. Unlike mutating joins, these do not add new columns to the outputted `tibble`. Instead, they filter rows in `x` depending on whether there is a match in `y` or not.

-   `semi_join(x, y, by = ...)`: Returns all rows from `x` that have matches in `y`. Keeps columns just from `x`.
-   `anti_join(x, y, by = ...)`: Returns all rows from `x` that **do not** have matches in `y`. Keeps columns just from `x`.

```{r}
semi_join(mcs_14y_mini,
          mcs_17y_mini,
          by = c("MCSID", FCNUM00 = "GCNUM00"))

anti_join(mcs_14y_mini,
          mcs_17y_mini,
          by = c("MCSID", FCNUM00 = "GCNUM00"))
```

Again, rows that are retained after a specific filtering `*_join()` imply a particular response pattern.

> **Question**: What response pattern does `semi_join()` imply? What about `anti_join()`?

### Appending `tibbles` with `dplyr::bind_rows()`

We have seen how the mutating `*_join()` functions can be used to merge `tibbles` together in wide format. Appending data with `bind_rows()` allows us to place data in long format. `bind_rows()` is simple - it takes a set of `tibbles` and then stacks these in the order they are specified. If the inputs are named, the `.id` argument can be used to populate a column in the outputted `tibble` that identifies the input `tibble` a given row came from. The input `tibbles` can be specified separately or combined in a list.

```{r}
# Two equivalent approaches.
# list(
#   `11` = mcs_11y_mini,
#   `14` = mcs_14y_mini,
#   `17` = mcs_17y_mini
# ) %>%
#   bind_rows(.id = "fup")

bind_rows(
  `11` = mcs_11y_mini,
  `14` = mcs_14y_mini,
  `17` = mcs_17y_mini, 
  .id = "fup"
)
```

`bind_rows()` retains all the columns from the input `tibbles`, even if a given column did not appear in all of them. We can see this create issues where column names differ across the input datasets. Therefore, variable names should be harmonised before piping into `bind_rows()`:

```{r}
list(
  `11` = mcs_11y_mini ,
  `14` = mcs_14y_mini,
  `17` = mcs_17y_mini
) %>%
  map(
    ~ .x %>% 
      rename_with(~ str_sub(.x, 2),
                  .cols = -MCSID)
  ) %>%
  bind_rows(.id = "fup")
```

Note, `bind_rows()` produces a warning where the column attributes vary between the input `tibbles`. Here the `value_labels` attributes varied. `bind_rows()` resolves inconsistencies by using the attributes from the first `tibble`.

This code has worked slightly better - it has placed values from Sweeps 6 & 7 in the same columns, but as Sweep 5 uses a slightly different rubric for variable names, these still appear separately. We could improve the code further with to handle this scenario (e.g., `str_sub(.x, 2. -3)`, but another helpful function is `dplyr::coalesce()`. `coalesce()` takes multiple objects as inputs, and at each position, uses the first non-missing value in the output.

```{r}
#| warning: false

list(
  `11` = mcs_11y_mini ,
  `14` = mcs_14y_mini,
  `17` = mcs_17y_mini
) %>%
  map_dfr(~ .x %>% rename_with(~ str_sub(.x, 2), .cols = -MCSID),
          .id = "fup") %>%
  mutate(height = coalesce(CHTCMA0, CHTCM00), 
         weight = coalesce(CWTCMA0, CWTCM00),
         .after = CNUM00)
```

Note, we use `map_dfr(.id = "fup")` here which achieves the same purpose as `map() %>% bind_rows(.id = "fup")`, but in a single function call.

Notice that `bind_rows()` outputs rows for all observations in the input `tibbles`, and implies a particular response pattern.

### Task III

1.  It is often more efficient to append raw data together and then clean the data. Amend the `clean_bmi()` code above to append the data from Sweeps 3-7 and then clean the output.

```{r}
#| output: false
#| code-fold: true
#| code-summary: "Reveal the solution"

load_anthro <- function(sweep){
  sweep_letter <- letters[sweep]
  fup <- mcs_fups[sweep]
  
  glue("{mcs_fld}/{fup}y/mcs{sweep}_cm_interview.dta") %>%
    read_dta(col_select = c("MCSID",
                            matches("^.(CNUM|C(W|H)TCM)(A|0)0"))) %>%
    rename(cnum = matches(".CNUM00"),
           height_raw = matches(".CHTCM(A|0)0"),
           weight_raw = matches(".CWTCM(A|0)0")) %>%
    mutate(sweep = !!sweep)
}

task3_solution <- map_dfr(3:7, load_anthro %>%
  mutate(cnum = as.integer(cnum),
         height = negative_to_na(height_raw) %>%
             as.numeric(),
           weight = negative_to_na(weight_raw) %>%
             as.numeric(),
           bmi = weight / ( (height / 100) ^ 2)) %>%
  select(MCSID, cnum, sweep, height, weight, bmi)

task3_solution
```

## Reshaping `tibbles` with `tidyr::pivot_*()`

A key aspect of LDA is moving data between long and wide formats (and vice versa). The `pivot_longer()` and `pivot_wider()` functions from `tidyr` can be used to do this.

### Wide-to-Long Transformations with `tidyr::pivot_longer()`

We will continue with the "mini" datasets we have been using so far. Let's start by creating a wide-dataset with data from sweeps 5-7 using `full_join()` and `reduce()`. As we use `full_join()`, this dataset contains data for all three cohort members. To simplify things slightly, we'll rename the variables from Sweep 5 (`E...A0`) to share the same suffix as Sweeps 6 & 7 (`00`).

```{r}
df_mini_wide <- list(mcs_11y_mini,
                     mcs_14y_mini, 
                     mcs_17y_mini) %>%
  reduce(
    ~ full_join(
      .x %>% rename(CNUM = matches("CNUM")),
      .y %>% rename(CNUM = matches("CNUM")) ,
      by = c("MCSID", "CNUM")
    )
  ) %>%
  rename_with(~ str_replace(.x, "A0$", "00"))

df_mini_wide
```

`df_mini_wide` has one row per cohort member. This `tibble` is not *tidy* as the variable *names* encode information about the sweep of assessment. To make the data tidy, we instead need one row per cohort member x sweep, with one column for height and weight each and with ID variables for person (`MCSID` & `CNUM`) and for sweep (to be created). The simplest `pivot_longer()` call will not achieve our aims, but it is worth seeing the syntax.

```{r}
df_mini_wide %>%
  pivot_longer(
    cols = -c(MCSID, CNUM), # Columns to pivot (everything but MCSID and CNUM)
    names_to = "variable",
    values_to = "value"
  )
```

This isn't tidy because there is now only one value column (`value`) when there should be two (i.e., for height and weight, separately) and the `variable` column contains information on the sweep of assessment and the variable type (height or weight). Instead, we need to use `pivot_longer()` to perform a more complex transformation where we extract the information on sweep from the variable name, placing this information in a new *name* column, and also use the rest of the variable name (`CHTCM00` or `CWTCM00`) to name the (two) new *value* columns.

To do this, we can use the `names_pattern` and `names_to` arguments of `pivot_longer()`. `names_pattern` separates the takes a regular expression to capture groups within the variable names. One group will capture the sweep information (e.g., "F" or "G") and the other will capture the variable type (e.g., `CHTCM00` or `CWTCM00`). `names_to` then specifies the names of these new columns, using the `.value` placeholder to specify the names will be used for the new value columns.[^05_data_preparation-3]

[^05_data_preparation-3]: Even more complex transformations can be achieved with more regular expression groups. See the [`pivot_longer()` vignette](https://tidyr.tidyverse.org/articles/pivot.html#longer) for more details.

```{r}
df_mini_wide %>%
  pivot_longer(
    cols = matches("^.C(W|H)TCM00"), # The columns to pivot.
    names_pattern = "(.)(.*)", # Two groups: first letter, and subsequent letters
    names_to = c("sweep", ".value")) # Names for the new groups
```

If we really insist, we can clean up the resulting variable names a bit more altering `names_pattern`.

```{r}
df_mini_wide %>%
  pivot_longer(
    cols = matches("^.C(W|H)TCM00"), # The columns to pivot.
    names_pattern = "(.)C(..)CM00", # Two groups: first letter, and third and fourth letters
    names_to = c("sweep_letter", ".value"))  %>%
  mutate(sweep = match(sweep_letter, LETTERS),
         fup = mcs_fups[sweep],
         .before = sweep_letter)
```

The resulting `tibble` has 9 rows - 3 for each cohort member (1 for each cohort member x sweep combination). This includes rows for sweeps a cohort member did not even appear in.

The above code shows how to reshape raw data, but the same principles apply to clean data to. With that in mind, when cleaning data, one should use consistent naming schemes for variables to make reshaping simpler. Using underscores (`_`) in variable names and placing sweep or follow-up age as a suffix is particularly helpful.

```{r}
rename_col <- function(var_name){
  var_dict <- c("CHTCM00" = "height", "CWTCM00" = "weight")
  sweep_letter <- str_sub(var_name, 1, 1)
  clean_var <- var_dict[str_sub(var_name, 2)]
  glue("{clean_var}_{mcs_fups[match(sweep_letter, LETTERS)]}")
}

df_mini_wide %>%
  mutate(across(where(is.labelled), negative_to_na)) %>%
  rename_with(rename_col, .cols = -c(MCSID, CNUM)) %>%
  pivot_longer(
    cols = -c(MCSID, CNUM),
    names_pattern = "(.*)_(.*)", 
    names_to = c(".value", "fup"),
    names_transform = list(fup = as.integer) # To change `fup` column from character to integer
  )
```

#### Task IV

1.  Change `names_to = c(".value", "sweep_letter")` in the code above and run. Before looking at the output, predict what the resulting dataset will look like. What will the columns names be? How many rows of data will there be in the output? Can you think of a purpose of data in this format? (Hint: consider could be achieved with `group_by()`).

```{r}
#| output: false
#| code-fold: true
#| code-summary: "Reveal the solution"

df_mini_wide %>%
  pivot_longer(
    cols = matches("^.C(W|H)TCM00"), # The columns to pivot.
    names_pattern = "(.)C(.*)CM00", # Two groups: first letter, and subsequent letters
    names_to = c(".value", "variable")) %>%
  group_by(variable) %>%
  summarise(rho = cor(E, G, use = "complete.obs"))
```

### Long-to-Wide Transformations with `tidyr::pivot_wider()`

Let's now consider the opposite transformation - moving data from long to wide format. `pivot_wider()` shares some syntax with `pivot_longer()` but there are also key differences. We'll again use the "mini" datasets and start with a simple function call, reshaping a single column, and then move on to reshaping multiple columns at once.

```{r}
#| warning: false

df_long_mini <- list(
  `11` = mcs_11y_mini %>% rename_with(~ str_replace(.x, "A0$", "00")),
  `14` = mcs_14y_mini,
  `17` = mcs_17y_mini
) %>%
  map_dfr(~ .x %>% rename_with(~ str_sub(.x, 2), .cols = -MCSID),
          .id = "fup") %>%
  mutate(fup = as.integer(fup)) %>%
  arrange(MCSID, CNUM00, fup) %>%
  relocate(fup, .after = CNUM00)

df_long_mini
```

Let's pivot the height (`CHTCM00`) variable wider. By specifying the `id_cols` argument, we can specify the columns to keep as is in the outputted `tibble` - as `CWTCM00` does not appear in the specified columns, it is thrown away.

```{r}
df_long_mini %>%
  pivot_wider(
    id_cols = c(MCSID, CNUM00),
    names_from = fup,
    values_from = CHTCM00
  )
```

> **Question:** Remove the `id_cols = ...` argument in the code above and run. What happens and why?

Note, the new column names in the output above are not very helpful as the entire name is taken from the `fup` column. Instead, the `names_glue()` argument can be used specify a rubric for the new column names, using `glue()`-like syntax. As with `pivot_longer()`, we can use the `.value` placeholder to use the name of the values column(s).

```{r}
df_long_mini %>%
  pivot_wider(
    id_cols = c(MCSID, CNUM00),
    names_from = fup,
    values_from = CHTCM00,
    names_glue = "{.value}_{fup}"
  )
```

Generalising this code to situations where we want to reshape multiple columns at once is easy - we can just amend the `values_from` argument.

```{r}
df_long_mini %>%
  pivot_wider(
    id_cols = c(MCSID, CNUM00),
    names_from = fup,
    values_from = matches("C(W|H)TCM00"),
    names_glue = "{.value}_{fup}",
    names_vary = "slowest" # Places variables from the same sweep side-by-side
  )
```

Notice in the output that we have one row per cohort member and a `NA` value for the columns for sweeps they may not have even responded in. `pivot_wider()` makes [implicit missing values explicit](https://r4ds.hadley.nz/missing-values.html).

#### Task V

1.  Take the output of the solution to Task III and reshape the weight and BMI columns to wide format, discarding height. Keep rows for only those individuals who appeared in all sweeps.

```{r}
#| output: false
#| code-fold: true
#| code-summary: "Reveal the solution"

task3_solution %>%
  add_count(MCSID, cnum) %>%
  filter(n == 5) %>%
  pivot_wider(
       id_cols = c(MCSID, cnum),
    names_from = sweep,
    values_from = c(weight, bmi),
    names_glue = "{.value}_{sweep}",
  )
```

## Some Helpful Functions for Cleaning Long Data

We'll finish with a few functions that are helpful when cleaning long data.

### `tidyr::complete()`

While `pivot_wider()` makes implicit values explicit, `bind_rows()` does not. It is possible to create missing rows when data are in long format by using `tidyr::complete()`. This function is straightforward - you pass it a `tibble` and then specify the columns one wishes to get every combination for (here the ID and time variables). By default, other columns will be filled with `NA` values in the newly created rows. This behaviour can be overridden using the `fill` argument. 

One issue with MCS data is that ID columns are spread over two variables. This creates issues if you have multiple values of `CNUM00` in the data. We can avoid the problem by first *uniting* (`unite()`) the ID columns then separating (`separate()`) them after using `complete()`.

```{r}
df_long_mini %>%
  unite("iid", MCSID, CNUM00, sep = "_") %>%
  complete(iid, fup) %>%
  separate(iid, 
           into = c("MCSID", "CNUM00"), 
           sep = "_",
           convert = TRUE)
```

### `tidyr::fill()`

`fill()` allows one to fill in missing values in a column with observed values appearing above or below. This is particularly useful when a time-invariant variable is only available for a person in some survey sweeps, or where such an assumption about time-invariance is justified. For instance, in tobacco research, the conservative Russell Standard ([West et al., 2005](https://doi.org/10.1111/j.1360-0443.2004.00995.x)) is sometimes used which assumes that smokers who drop-out of a survey continue to smoke. Another example is adult height, which is often only collected from those for whom data is not already available (e..g, as in the NCDS), and can be assumed to be stable over adulthood. Available measures can therefore be used to deterministically impute the missing values.

The `fill()` function is straightforward. One passes it a `tibble` (possibly grouped) and the columns to fill, along with the direction to fill values (using argument `.direction`): down, up, downup, or updown. Below we use `complete()` to make missing sweeps explicit, then fill the missing height values with the last observed one for that individuals.

```{r}
df_long_mini %>%
  complete(MCSID, CNUM00, fup) %>% # Simplified as CNUM == 1 always
  group_by(MCSID, CNUM00) %>%
  fill(CHTCM00, .direction = "down") %>% 
  ungroup()
```

Of course, this assumption of invariance in height is invalid here, given the data are from adolescence, not adulthood.

### Some `dplyr` Helpers

`dplyr` contains several helpful functions for working with grouped data:

-   `lag()` and `lead()`: Returns the preceding and succeeding values in an object (e.g., a column vector), respectively
-   `first()` & `last()`: Returns the first and last values in an object, respectively.
-   `nth()`: Returns the nth value in an object
-   `n_distinct()`: Returns the number of distinct values in an object.
-   `n()`: Returns the number of rows in the current group.
-   `row_number()`: Returns the row number within each group.

```{r}
df_long_mini %>%
  select(-CHTCM00, -CWTCM00) %>%
  group_by(MCSID, CNUM00) %>%
  mutate(
    prev_fup = lag(fup), 
    next_fup = lead(fup),
    observation = row_number()
  ) %>%
  ungroup()ZA`

df_long_mini %>%
  group_by(MCSID, CNUM00) %>%
  summarise(
    first_fup = first(fup), 
    last_fup = last(fup), 
    second_fup = nth(fup, 2),
    n_observations = n(), 
    n_distinct_fups = n_distinct(fup),
    .groups = "drop"
  )
```

## Putting It All Together

XX-MENTION INTERLUDE
